---
title: "Average Age of Death between Marital Status (For Year 2015)"
author: "Steven Seeger, Seth Pixton"
date: "12/06/2018"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
output: pdf_document
---

```{r setup, echo=FALSE}
library(knitr)
library(kableExtra)
library(invgamma)
load("data.RData")
load("Summary Table.RData")

if(!exists("married")&!exists("divorced")) source("load Data.R")
if(!exists("divorcedMu")) source("Creating the Posterior(Divorced).R")
if(!exists("marriedMu")) source("Creating the Posterior(Married).R")
```


# Introduction

The question we decided to research is this: Does the marital status of an individual in the United States affect the average age of death? The marital statuses we will be looking at are Divorced and Married. The value of this question is that if we can determine that marital status has an effect on average age of death, it could influence the insurance industry and encourage further research as to why this is the case.

All four groups' EDA are included for the purpose of viewing and later interest. We will not be exploring this further than that.

# Literature Review

A study was conducted by Dustin C. Brown about Life Expectancy Differentials by Marital Status, Individuals’ Own Education, and Spousal Education in the United States. (http://paa2014.princeton.edu/papers/142823) This study was interesting, but did not exactly match the study that we are conducting. He used life expectancy as his research and we are using average age of death. His conclusion was that there is a difference between many groups for life expectancy at age 55. We are looking at all people that are married at any age and all that are married and then divorced.

# Methods

We gathered our data from Kaggle.com. (Cite: https://www.kaggle.com/cdc/mortality/home) From the documentation for this data set, it states that "[a]ll data comes from the CDC’s National Vital Statistics Systems, with the exception of the Icd10Code, which are sourced from the World Health Organization." We are curious about the mean age of death for people who are married versusthat of people who are divorced. We choose to go with a Normal-Normal-Inverse-Gamma prior distribution. We chose this distribution because we decided that deaths will probably be normally spread out around the center of the data. We went with an uninformative prior of mu ~ N(0, 10^2) and sigma^2 ~ IG(0.01, 0.01). We felt that this was appropriate given the vast amount of data we had.

The following two graphs show the Prior Distributions we decided to use.

```{r}
df <- rnorm(10000, 0, 10)

df <- as.data.frame(df)

ggplot(df, aes(x=df)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", binwidth = 1)+
 geom_density(alpha=.2, fill="#FF6666")+
  ggtitle("Prior N(0,10^2)")

x <- seq(0, 5, .01)
qplot(x, dinvgamma(x, 0.01, 0.01), geom = "line") +
  ggtitle("Prior IG(0.01,0.01)")
```

The posterior distribtuions will answer our research question by allowing us to see if our data is different at a significant level. We can use the difference in means in the posterior distributions to detect if there is a true difference between the age of death between the two groups.

# Exploratory Data Analysis

Included in the following section is a summary of the data, as well as density distributions of the data. The first chart that will be shown is a boxplot. We are able to include a boxplot in this report because the extreme outliers were removed, and the chart looks normal and is easier to understand.

\blandscape

```{r label1, out.width = "100%", fig.cap = "Boxplot", echo = FALSE}
include_graphics("Report Files/Boxplot.pdf")
```

\elandscape

Shown in the table below, are the summary statistics for each of the groups we were testing.

```{r SummaryStats, echo=FALSE}
kable(TableSummary, caption = "Summary Statistics (age of death)")%>%
kable_styling(latex_options = c("striped", "hold_position", "scale_down"))
```

The mean, median and standard deviation for the groups are shown. The mean and the median for each group are almost identical.

Shown on the next page are several density distribution charts. This helped us determine which of the two groups to compare.

\newpage
\blandscape

```{r label, out.width = "100%", fig.cap = "Density Plots", echo = FALSE}
include_graphics("Report Files/Density Plot.pdf")
```

\elandscape

# Results
```{r}
diff<-marriedMu[-c(1:500)]-divorcedMu[-c(1:500)]
mean(diff)
quantile(diff,c(0.025,0.975))
```

# Discussion
Our research used the data tha only include each individual marital status, and the age of death. We did not have the age at which the individuls married or divorced, or how many times they were married. Additionally, our data only included individuals in the United States. Further research can be conducted that will take into account these variables, but we are able to show that on average, married individuals live over 3 years longer than divorced individuals.